**NeuroKit2: A Python Toolbox for Neurophysiological Signal Processing**

### 2020-09-08 paper/manuscript_v3.Rmd

---
title: '**NeuroKit2: A Python Toolbox for Neurophysiological Signal Processing**'
shorttitle        : "NeuroKit2"
author:
  - name          : "Dominique Makowski"
    affiliation   : " 1,*"
    corresponding : no    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dmakowski@ntu.edu.sg"
  - name          : "Tam Pham"
    affiliation   : " 1"
  - name          : "Zen J. Lau"
    affiliation   : " 1"
  - name          : "Jan C. Brammer"
    affiliation   : " 2"
  - name          : "François Lespinasse"
    affiliation   : " 3, 4"
  - name          : "Hung Pham"
    affiliation   : " 5"
  - name          : "Christopher Schölzel"
    affiliation   : " 6"
  - name          : "S.H. Annabel Chen"
    affiliation   : " 1, 7, 8"
affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "Behavioural Science Institute, Radboud University, Netherlands"
  - id            : "3"
    institution   : "Département de psychologie, Université de Montréal, Canada"
  - id            : "4"
    institution   : "Centre de Recherche de l'Institut Universitaire Geriatrique de Montréal"
  - id            : "5"
    institution   : "Eureka Robotics, Singapore"
  - id            : "6"
    institution   : "Life Science Informatics, THM University of Applied Sciences, Germany"
  - id            : "7"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
  - id            : "8"
    institution   : "Lee Kong Chian School of Medicine, Nanyang Technological University, Singapore"
authornote: |
  \* Correspondence concerning this article should be addressed to Dominique Makowski (HSS 04-18, 48 Nanyang Avenue, Singapore; dmakowski@ntu.edu.sg).
abstract: |
  NeuroKit2 is an open-source, community-driven, and user-friendly Python package dedicated to neurophysiological signal processing with an initial focus on bodily signals (e.g., ECG, PPG, EDA, EMG, RSP). Its design philosophy is centred on user-experience and accessibility to both novice and advanced users. The package provides a consistent set of high-level functions that enable data processing in a few lines of code using validated pipelines, which we illustrate in two examples covering the most typical scenarios, such as an event-related paradigm and an interval-related analysis. The package also includes tools dedicated to specific processing steps such as rate extraction and filtering methods, offering a trade-off between efficiency and fine-tuned control to the user. Rather than focusing on specific signals, NeuroKit2 was developed to provide a comprehensive means for a simultaneous processing of a wide range of signals. Its goal is to improve transparency and reproducibility in neurophysiological research, as well as foster exploration and innovation.
keywords          : "Neurophysiology, Biosignals, Python, ECG, EDA, EMG"
wordcount         : "2513"
bibliography      : ["bibliography.bib"]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output:
  papaja::apa6_pdf:
    keep_tex: FALSE
    latex_engine: xelatex
  papaja::apa6_word:
    keep_tex: FALSE
header-includes:
   - \usepackage[labelfont=bf, font={color=gray,small}]{caption}
   - \usepackage{float}
   - \usepackage[document]{ragged2e}
editor_options:
  chunk_output_type: console
---

@@

<!-- Research gap 1 -->
Neurophysiological measurements increasingly gain popularity in the study of cognition and behavior.
These measurements include electroencephalography (EEG), electrocardiography (ECG), electromyography (EMG) and electrodermal activity (EDA).
Their popularity is driven by theoretical motivations [e.g., the growth of embodied or affective neuroscience; @Kiverstein2015] as well as practical reasons.
The latter include low costs (especially compared with other imaging techniques, such as MRI or MEG), ease of use (e.g., portability, setup speed), and the increasing availability of recording devices [e.g., wearables; @yuehong2016internet].
Moreover, the extraction of meaningful information from neurophysiological signals is facilitated by current advances in signal processing algorithms [@clifton2012machine; @roy2019deep].
Unfortunately, these algorithms are mostly inaccessible to researchers without experience in programming and signal processing.
Moreover, many software tools for neurophysiological analyses are limited to a single type of signal (for instance, focused on ECG).
This makes it inconvenient for researchers who might have to learn and concurrently rely on different software to process multimodal data.

<!-- Research gap 2 -->
Additionally, psychology and neuroscience face a "reproducibility crisis" [@topalidou2015long; @open2015estimating; @milkowski2018replicability; @maizey2019barriers] which has lead to a profound reassessment of research practices (by researchers, publishers, fund agencies, ...).
The opacity of data processing, such as ill-specified, or incaccessible analysis pipelines, plays a major role in the crisis.
This issue could in part be alleviated by making analyses code an integral part of scientific publications, rather than treating a paper as the sole and most important part of the research  project. However, distributing the analysis script alongside the paper poses new challenge:
Scripts must be shareable (not always feasible with closed-source and proprietary software or programming languages), accessible (well-documented and organized scripts) and reproducible (which is inherently difficult for software relying on a graphical user interfaces - GUI - in which the manual point-and-click sequence is hard to automate).

<!-- Response to gap -->
*NeuroKit2* addresses these challenges by offering a free, user-friendly, and comprehensive solution for neurophysiological data processing, with an initial focus on bodily signals (including ECG, PPG, RSP, EDA, EMG, EOG) and generic functions that could also support other signal processing such as EEG (for which more specific support is in development).
It is an open-source Python package, developed by a multi-disciplinary team that actively invites new collaborators.
It aims at being accessible, well-documented, well-tested, cutting-edge, flexible and efficient, allowing users to select from a wide range of validated analysis pipelines as well as creating their own.
Historically, *NeuroKit2* is the re-forged successor *NeuroKit1* [@neurokit1], taking on its most successful features and design choices, and re-implementing them adhering to best practices in research software development.

<!-- Accessibility and documentation -->
The package is implemented in Python 3 [@python3], which means that *NeuroKit2's* users benefit from an large number of learning resources and a vibrant community. The package depends on relatively few, well established and robust packages from the Python data analysis ecosystem [@scipy] such as *NumPy*, *pandas*, *SciPy*, *scikit-learn* and *MatplotLib* (with an additional system of optional dependencies), making *NeuroKit2* itself a viable dependency in other software.

@@

As stated above, *NeuroKit2* aims at being accessible to beginners and, at the same time, offering a maximal level of control to experienced users. This is achieved by allowing beginning users to implement complex processing and analyses pipelines with very few functions, while still enabling fine-tuned control over arguments and parameters to more experienced users. In concrete terms, this trade-off is enabled by an API structure organized in three three layers.

@@

The mid-level functions are assembled in high-level wrappers, that are convenient entry points for new users. For instance, the `ecg_process()` function internally chains the mid-level functions `ecg_clean()`, `ecg_findpeaks()`, `ecg_rate()`. A specific processing pipeline can be selected with the `method` argument that is then propagated throughout the internal functions. Easily switching between processing pipelines allows for the comparison of different methods, and streamlines critical but time-consuming steps in reproducible research, such as the validation of data preparation and quality control [@Quintana2016]. Finally, the package includes convenience meta-functions (e.g., `bio_process`) that enable the combined processing of multiple types of signals at once (e.g., `bio_process(ecg=ecg_signal, eda=eda_signal)`).

Performing an entire set of operations with sensible default parameters in one function can be rewarding, especially for beginners, allowing them to perform cutting-edge processing or replication of research steps without requiring much programming expertise. Moreover, it contributes to the demystification of the usage of programming tools (as opposed to GUI-based software such as *SPSS*, *Kubios*, or *Acqknowledge*), providing a welcoming framework to further explore the complexities of physiological data processing. Importantly, more advanced users can easily build custom analysis pipelines by using the low- and mid-level functions, allowing for a finer control over the processing parameters. We believe that this implementation is a well-calibrated trade-off between flexibility and user-friendliness.

@@

This example dataset contains ECG, RSP and EDA signals of one participant who was presented with four emotional images [from the NAPS database; @marchewka2014nencki] in a typical (albeit highly shortened) experimental psychology paradigm.

The signals are 2.5 minutes long and are recorded at a frequency of 100Hz [note that the sampling rate is low for storage purposes and should be higher in actual recordings, see @Quintana2016]. It has 4 channels including three physiological signals, and one corresponding to events marked with a photosensor (signal strength decreases when a stimulus appeared on the screen).

@@

In this example, the steps of the analysis are identical to the previous example, including loading the package, the dataset and processing the data. Furthermore, Stimulus onsets in the photosensor are detected separately with `events_find()`. Once we have the preprocessed signals and the location of events, we can slice the data into segments corresponding to a time window (ranging from -0.1 to 4 seconds) around each stimulus with `epochs_create()`. Finally, relevant features are computed for each epoch (i.e., each stimulus) by passing them to `bio_analyze()`.

@@

This example shows the straightforward process of extracting features of physiological responses. This pipeline can easily scale up to group-level analysis by aggregating the average of features across participants. In addition to streamlining data analyses, *NeuroKit2* aims to provide researchers with an extensive suite of signal features, allowing for precise interpretations in terms of relationship between physiological activity and neurocognitive processes. In this example (see **Table 2**), exposure to negative stimuli, as compared to neutral stimuli, is related to stronger cardiac deceleration, higher skin conductance response, and accelerated breathing rate (note that this descriptive interpretation is given solely for illustrative purposes).

@@

*NeuroKit2* is a neurophysiological signal processing library accessible to people with all levels of programming experience and backgrounds. Its development is focused on creating an intuitive user-experience, as well as building a collaborative community. It is also a pragmatic answer to the broader need for transparent and reproducible methods in neurophysiology. Its modular structure and organization not only facilitates the use of existing and validated processing pipelines, but also creates a fertile ground for experimentation and innovation.

We expect the package's future evolution to be driven by the communities' needs and the advances in related fields. For instance, although *NeuroKit2* already implements a lot of useful functions for EEG processing (such as entropy and fractal dimensions quantification), its support could be further improved [for example with high-level functions built on top of utilities provided by the leading EEG Python software, namely *MNE*; @gramfort2013meg]. Additionally, in the future we strive to support other types of bodily signals (e.g., electrogastrography - EGG, electrooculography - EOG) and plan to optimize computational efficiency on large datasets. We also plan to further validate the available processing pipelines using public databases. In line with this objective, the support of standardized data structure formats (e.g. WFDB, BIDS, ...) could be extended.

In conclusion, we believe that *NeuroKit2* provides useful tools for anyone who is interested in analyzing physiological data collected with research-grade hardware or wearable "smart health devices". By increasing the autonomy of researchers and practitioners, and by shortening the delay between data collection and results acquisition, *NeuroKit2* could be useful beyond academic research in neuroscience and psychology, including applications such as personal physiological monitoring and exercise science. Finally, we hope that *NeuroKit2* encourages users to become part of a supportive open-science community with diverse areas of expertise rather than relying on closed-source and proprietary software, thus shaping the future of neurophysiology and its related fields.

>>>>

---
title: '**NeuroKit2: A Python Toolbox for Neurophysiological Signal Processing**'
shorttitle        : "NeuroKit2"
author:
  - name          : "Dominique Makowski"
    affiliation   : " 1,*"
    corresponding : no    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dmakowski@ntu.edu.sg"
  - name          : "Tam Pham"
    affiliation   : " 1"
  - name          : "Zen J. Lau"
    affiliation   : " 1"
  - name          : "Jan C. Brammer"
    affiliation   : " 2"
  - name          : "François Lespinasse"
    affiliation   : " 3, 4"
  - name          : "Hung Pham"
    affiliation   : " 5"
  - name          : "Christopher Schölzel"
    affiliation   : " 6"
  - name          : "S.H. Annabel Chen"
    affiliation   : " 1, 7, 8"
affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "Behavioural Science Institute, Radboud University, Netherlands"
  - id            : "3"
    institution   : "Département de psychologie, Université de Montréal, Canada"
  - id            : "4"
    institution   : "Centre de Recherche de l'Institut Universitaire Geriatrique de Montréal"
  - id            : "5"
    institution   : "Eureka Robotics, Singapore"
  - id            : "6"
    institution   : "Life Science Informatics, THM University of Applied Sciences, Germany"
  - id            : "7"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
  - id            : "8"
    institution   : "Lee Kong Chian School of Medicine, Nanyang Technological University, Singapore"
authornote: |
  \* Correspondence concerning this article should be addressed to Dominique Makowski (HSS 04-18, 48 Nanyang Avenue, Singapore; dmakowski@ntu.edu.sg).
abstract: |
  NeuroKit2 is an open-source, community-driven, and user-friendly Python package dedicated to neurophysiological signal processing with an initial focus on bodily signals (e.g., ECG, EDA, EMG, EOG, PPG etc.). Its design philosophy is centred on user-experience and accessibility to both novice and advanced users. The package provides a consistent set of high-level functions that enable data processing in a few lines of code using validated pipelines, which we illustrate in two examples covering the most typical scenarios, such as an event-related paradigm and an interval-related analysis. The package also includes tools dedicated to specific processing steps such as rate extraction and filtering methods, offering a trade-off between efficiency and fine-tuned control to the user. Rather than focusing on specific signals, NeuroKit2 was developed to provide a comprehensive means for a simultaneous processing of a wide range of signals. Its goal is to improve transparency and reproducibility in neurophysiological research, as well as foster exploration and innovation.
keywords          : "Neurophysiology, Biosignals, Python, ECG, EDA, EMG"
wordcount         : "2513"
bibliography      : ["bibliography.bib"]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output:
  papaja::apa6_pdf:
    keep_tex: FALSE
    latex_engine: xelatex
  papaja::apa6_word:
    keep_tex: FALSE
header-includes:
   - \usepackage[labelfont=bf, font={color=gray,small}]{caption}
   - \usepackage{float}
   - \usepackage[document]{ragged2e}
editor_options:
  chunk_output_type: console
---

@@

<!-- Research gap 1 -->
Neurophysiological measurements increasingly gain popularity in the study of cognition and behavior.
These measurements include electroencephalography (EEG), electrocardiography (ECG), electromyography (EMG) and electrodermal activity (EDA).
Their popularity is driven by theoretical motivations [e.g., the growth of embodied or affective neuroscience; @Kiverstein2015] as well as practical reasons.
The latter include low costs (especially compared with other imaging techniques, such as MRI or MEG), ease of use (e.g., portability, setup speed), and the increasing availability of recording devices [e.g., wearables; @yuehong2016internet].
Moreover, the extraction of meaningful information from neurophysiological signals is facilitated by current advances in signal processing algorithms [@clifton2012machine; @roy2019deep].
Unfortunately, these algorithms are mostly inaccessible to researchers without experience in programming and signal processing.
Moreover, many software tools for neurophysiological analyses are limited to one type of signal (for instance, focused on ECG).
This makes it inconvenient for researchers who might have to learn and concurrently rely on different software to process multimodal data.

<!-- Research gap 2 -->
Another important issue existing in psychology and neuroscience has been coined as the "reproducibility crisis" [@topalidou2015long; @open2015estimating; @milkowski2018replicability; @maizey2019barriers], and has lead to a profound questioning and reassessment from different actors involved (researchers, publishers, fund agencies, ...).
One of the main identified contributing factor is the actual opacity of data processing, where analysis pipelines are not described in enough details to ensure a full and exact reproduction.
One of the suggested response to that issue has been to provide, alongside the study, the analysis script, which in turns opens new challenges.
Indeed, these scripts must be shareable (not always feasible with closed-source and proprietary software or programming languages), accessible (enticing documented and well-organized scripts) and reproducible (which is inherently difficult for many software relying on a graphical user interface - GUI - in which the manual point-and-click sequence is hard to automate).

<!-- Response to gap -->
*NeuroKit2* addresses these challenges by offering a free, user-friendly, and comprehensive solution for neurophysiological data processing, with an initial focus on bodily signals (including ECG, PPG, RSP, EDA, EMG, EOG) and generic functions that could also support other signal processing such as EEG (for which more specific support is in development).
It is an open-source Python package, developed by a multi-disciplinary team that actively invites new collaborators.
It aims at being accessible, well-documented, well-tested, cutting-edge, flexible and efficient, allowing users to select from a wide range of validated analysis pipelines as well as creating their own.
Historically, *NeuroKit2* is the re-forged successor *NeuroKit1* [@neurokit1], taking on its most successful features and design choices, and re-implementing them in a professional and well-thought way.

<!-- Accessibility and documentation -->
The package is implemented in Python 3 [@python3], which means that *NeuroKit2's* users benefit from an large amount of learning resources and a vibrant community. The package depends on relatively few, well established and robust packages from the Python data analysis ecosystem [@scipy] such as *NumPy*, *pandas*, *SciPy*, *scikit-learn* and *MatplotLib* (with an additional system of optional dependencies), making *NeuroKit2* itself a viable dependency in other software.

@@

As stated above, *NeuroKit2* aims at being accessible to beginners and, at the same time, offering a maximal level of control to experienced users. This is achieved by allowing beginning users to implement complex processing and analyses pipelines with very few functions, while still enabling fine-tuned control and precision over arguments and parameters to more experienced users. In concrete terms, this trade-off is allowed by a API structure organized in three three layers of abstraction.

@@

The mid-level functions are assembled in high-level "master" functions, that are convenient entry points for new users. For instance, the `ecg_process()` function internally chains the mid-level functions `ecg_clean()`, `ecg_findpeaks()`, `ecg_rate()`. A specific processing pipeline can be selected with the `method` argument, that is then propagated throughout the internal functions. Easily switching between processing pipelines allows for the comparison of different methods, and streamlines critical but time-consuming steps in reproducible research, such as the validation of data preparation and quality control [@Quintana2016]. Finally, the package includes convenience meta-functions (e.g., `bio_process`) that enable the combined processing of multiple types of signals at once (e.g., `bio_process(ecg=ecg_signal, eda=eda_signal)`).

Performing an entire set of operations with sensible default parameters in one function can be rewarding, especially for beginners, allowing them to perform cutting-edge processing or replication of research steps without requiring much programming expertise. Moreover, it contributes to the demystification of the usage of "pure" programming tools (as opposed to GUI-based software such as *SPSS*, *Kubios*, or *Acqknowledge*), providing a welcoming framework to further explore the complexities of physiological data processing. Importantly, more advanced users can easily build custom analysis pipelines by using the mid-level functions, allowing for a finer control over the processing parameters. We believe that this implementation is a well-calibrated trade-off between flexibility and user-friendliness.

@@

This example dataset contains ECG, RSP and EDA signals of one participant who was presented with four emotional images [from the NAPS database; @marchewka2014nencki], in a typical (albeit highly shortened) experimental psychology paradigm.

Signals are 2.5 minutes long and are recorded at a frequency of 100Hz [note that the sampling rate is low for storage purposes and should be higher in actual recordings, see @Quintana2016]. It has 4 channels including three physiological signals, and one corresponding to the marking of events with a photosensor (which signal decreases when a stimulus appeared on the screen).

@@

In this example, the steps of the analysis are identical to the precedent, including loading the package, the dataset and processing the data. Furthermore, Stimulus onsets in the photosensor are detected separately with `events_find()`. Once we have the preprocessed signals and the location of events, we can slice the data into segments corresponding to a time window (here, ranging from -0.1 to 4 seconds) around each stimulus with `epochs_create()`. Finally, relevant features are computed for each epoch (i.e., each stimulus) by passing them to `bio_analyze()`.

@@

This example shows the straightforward process of extracting features of physiological responses. This pipeline can easily scale up to group-level analysis by aggregating the average of features across participants. In addition to streamlining data analyses, *NeuroKit2* aims to provide researchers an extensive suite of signal features, allowing for precise interpretations in terms of relationship between physiological activity and neurocognitive processes. In this example (see **Table 2**), exposure to negative stimuli, as compared to neutral stimuli, is related to stronger cardiac deceleration, higher skin conductance response, and accelerated breathing rate (note that this descriptive interpretation is given solely for illustrative purposes).

@@

*NeuroKit2* is a neurophysiological signal processing software accessible to people with all levels of programming experience and background. Its development is focused on creating an intuitive user-experience, as well as building a collaborative community. It is also a pragmatic answer to the broader need for transparent and reproducible methods in neurophysiology. Its modular structure and organization not only facilitates the use of existing and validated processing pipelines, but also creates a fertile ground for experimentation and innovation.

We expect the package's future evolution to be driven by the communities' needs and the advances in related fields. For instance, although *NeuroKit2* already implements a lot of useful functions for EEG processing (such as entropy and fractal dimensions quantification), its support could be further improved [for example with high-level functions built on top of utilities provided by the leading EEG Python software, namely *MNE*; @gramfort2013meg]. Possible other future directions include extending the support for other types of bodily signals (e.g., electrogastrography - EGG, electrooculography - EOG) and achieving performance gains for large datasets by using efficient algorithms. Further validation of the available processing pipelines could be made through the (re)analysis of public databases. In line with this objective, the support of standardized data structure formats (e.g. WFDB, BIDS, ...) could be extended.

In conclusion, we believe that *NeuroKit2* provides useful tools for anyone who is interested in analyzing physiological data from research-grade hardware as well as wearable "smart health devices". By increasing the autonomy of researchers and practitioners, and by shortening the delay between data collection and results acquisition, *NeuroKit2* could be useful beyond academic research in neuroscience and psychology, including applications such as biofeedback, personal physiological monitoring and exercise science. Finally, we hope that *NeuroKit2* encourages users to become part of a supportive open-science community with diverse areas of expertise rather than relying on closed-source and proprietary software, thus shaping the future of neurophysiology and its related fields.
